---
title: "The gRain package"
output: html_notebook
---

### Installation

See <https://people.math.aau.dk/~sorenh/software/gR/#Packages> and for explanations of the chest example see https://www.norsys.com/tutorials/netica/secA/tut_A2.htm. 


```{r eval=FALSE, include=FALSE}
# Run only once, if/when needed! 
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("graph")
BiocManager::install("Rgraphviz")
BiocManager::install("RBGL") # added because needed it seems PR
```

Required packages:

```{r include=FALSE}
library(graph)
library(Rgraphviz)
library(RBGL)
library(gRbase)
library(gRain)
```

# The Chest Clinic example

Lauritzen and Spiegelhalter (1988) motivate the chest clinic example with the following narrative:

“Shortness–of–breath (dyspnoea) may be due to tuberculosis, lung cancer or bronchitis, or none of them, or more than one of them. A recent visit to Asia increases the chances of tuberculosis, while smoking is known to be a risk factor for both lung cancer and bronchitis. The results of a single chest X–ray do not discriminate between lung cancer and tuberculosis, as neither does the presence or absence of dyspnoea.”


-   Dyspnoea is the presented symptom.
-   Diagnosis can be
    -   tuberculosis, lung cancer or bronchitis,
    -   or a combination of them,
    -   or none of them.
-   Tuberculosis is made more likely if a recent visit to Asia.
-   Lung cancer and bronchitis are more likley in smokers
-   A chest X-ray does not discriminate between lung cancer and TB
-   Dysponea does not discriminate between lung cancer and TB.

To diagnose a patient, values could be entered for some of nodes when they are known. This would allow us to re-calculate the probabilities for all the other nodes. Thus if we take a chest x-ray and the x-ray is abnormal, then the chances of the patient having TB or lung-cancer rise. If we further learn that our patient visited Asia, then the chances that they have tuberculosis would rise further, and of lung-cancer would drop (since the X-ray is now better explained by the presence of TB than of lung-cancer). We will see how this is done in a later section.

For more see e.g. <https://www.norsys.com/tutorials/netica/secA/tut_A1.htm#WhatIsABayesNet>

## Specifying the network

### Step 1: CPTs

One way to do this is by specifying Conditional Probability Tables (CPTs):

```{r}
yn <- c("yes", "no")
a <- cptable(~asia, values=c(1, 99), levels=yn)
t.a <- cptable(~tub|asia, values=c(5, 95, 1, 99), levels=yn)
s <- cptable(~smoke, values=c(5, 5), levels=yn)
l.s <- cptable(~lung|smoke, values=c(1, 9, 1, 99), levels=yn)
b.s <- cptable(~bronc|smoke, values=c(6, 4, 3, 7), levels=yn)
e.lt <- cptable(~either|lung:tub, values=c(1, 0, 1, 0, 1, 0, 0, 1), levels=yn)
x.e <- cptable(~xray|either, values=c(98, 2, 5, 95), levels=c("Abnormal", "Normal"))
d.be <- cptable(~dysp|bronc:either, values=c(9, 1, 7, 3, 8, 2, 1, 9), levels=yn)
```

### Step 2: Compilation

Then compile the individual CPTs into a list

```{r}
chest_cpt <- compileCPT(a, t.a, s, l.s, b.s, e.lt, x.e, d.be)
chest_cpt
```

The components are arrays:

```{r}
chest_cpt$tub
```

Coercing into a datatable is sometimes easier to read:

```{r}
chest_cpt$tub  |> as.data.frame.table()
# using the pipe operator
```

Note that `either` is what's called a logical node:

```{r}
chest_cpt$either  |> as.data.frame.table()
```

### Step 3: Creating the BN

```{r}
chest_bn <- grain(chest_cpt) 
chest_bn
# the plot command may need to be run twice in interactive use
# the dag is an igraph object that prints nicer. 
plot(chest_bn$dag)
```

## Setting evidence and querying the BN

### Before adding evidence

Marginal probabilities is default:

```{r}
querygrain(chest_bn, nodes=c("lung", "bronc"), type="marginal")
```

We see from the baseline that bronchitis is much more frequent than lung cancer.

Joint probabilities on demand:

```{r}
querygrain(chest_bn, nodes=c("lung", "bronc"), type="joint")
```

```{r}
querygrain(chest_bn, nodes=c("either", "xray", "dysp"), type="marginal")
```

### Adding evidence

Evidence can be entered in various ways, here's one:

```{r}
chest_ev  <- setEvidence(chest_bn,
                evidence=list(asia="yes", dysp="yes"))
# display the evidence as a table:
getEvidence(chest_ev) |> as.data.frame()
```

Let's see the effect this has on the probs for the diagnosis lung cancer and bronchitis:

```{r}
qgrain(chest_ev, nodes=c("lung", "bronc"))
```

Bronchitis became much more likely.

We can also turn this around and ask how likely this evidence is (in absolute terms) under the model, kind of the evidence's base rate:

```{r}
pEvidence(chest_ev)
```

So, thats about 5 in thousand.

Let's save chest_ev to a file 

```{r}
saveRDS(chest_ev, file = "chest_ev.rds")
```

Now retract one evidence

```{r}
chest_ev_r <- retractEvidence(chest_ev, nodes = "dysp")
getEvidence(chest_ev_r)
```
Let's do this with chest_ev itself: 

```{r}
chest_ev <- retractEvidence(chest_ev, nodes = "dysp")
getEvidence(chest_ev)
```
Let's reload the saved model with the two evidences:

```{r}
chest_ev <- bn <<- readRDS("chest_ev.rds")
getEvidence(chest_ev)
```


### More examples for updates with evidence

https://www.norsys.com/tutorials/netica/secA/tut_A3.htm 

#### Observing Dyspnonea

```{r}
dysp_ev <- setEvidence(chest_bn,
  evidence=list( dysp="yes"))
# getFinding(dysp_ev)
```

Before and after the observation: 
```{r}
qgrain(chest_bn, nodes=c("lung", "bronc"))
qgrain(dysp_ev, nodes=c("lung", "bronc"))
```
#### Observing X-ray abnormal

```{r}
xRay_ev <- setEvidence(chest_bn,
  evidence=list(xray="Abnormal"))
```

Before and after the observation: 
```{r}
qgrain(chest_bn, nodes=c("lung", "bronc"))
qgrain(xRay_ev, nodes=c("lung", "bronc"))
```
xray being abnormal has increased the probability of lung cancer to 48.9%, i.e., from 5 out of hundred to 50 out of hundred. 

```{r}
getEvidence(xRay_ev)
```


```{r}
xRah_ev <- retractEvidence(xRay_ev, nodes = "xray")
```

```{r}
getEvidence(xRay_ev)
```


#### Explaining away
If we now here that she's been to Asia recently, the probability of cancer decreases to 37.1%  because a part of it is explained away by an increase of the chance to TB.

```{r}
xRay_Asia_ev <- setEvidence(xRay_ev,
  evidence=list(asia="yes"))
qgrain(xRay_ev, nodes=c("lung", "bronc"))
qgrain(xRay_Asia_ev, nodes=c("lung", "bronc"))
```
The general pattern for "explaining away" is as follows: 


Before we have any evidence for C, A and B are independent. That is, changing one has no affect on the other. But as soon as we have evidence for C, any change in the probability of A has an opposite change in B and vice versa. They are competing explanations of C. This is because if we acquire evidence of A, then that explains away C, and the chances of B are diminished. This is an important feature of good causal reasoning. It follows naturally from the laws of Bayesian statistics.

For serial blocking and divergent blocking see https://www.norsys.com/tutorials/netica/secA/tut_A3.htm

## Soft Evidence

A smaller net:

```{r}
yn <- c("yes", "no")
a    <- cpt(~asia, values=c(1, 99), levels=yn)
t.a  <- cpt(~tub|asia, values=c(5, 95, 1, 99), levels=yn)
plist1 <- compileCPT(list(a, t.a)) 
chest1 <- grain(plist1)
querygrain(chest1, simplify = TRUE)
```

### Hard evidence

```{r}
setEvidence(chest1, evidence=list(asia="yes"))
```

### Virtual (aka soft) evidence

We don't have to extend the network by an additional node, but add the non-binary evidence directly, with `asia=c(.8, .1)`: 

```{r}
chest1_ve <- chest1 |> setEvidence(evidence=list(asia=c(.8, .1)))
chest1_ve |> querygrain(simplify = TRUE)
```
The evidence is now 'soft': 

```{r}
getEvidence(chest1_ve, short=FALSE)
```

```{r}
querygrain(chest1, evidence=list(asia=c(1, 0)), simplify=T)
```

# Building networks from data

```{r}
dG <- dag(~A:B + B:C, result="igraph") 
uG <- ug(~A:B + B:C, result="igraph") 
par(mfrow=c(1,2)); plot( dG ); plot( uG )
```
Data:

```{r}
dat <- tabNew(c("A", "B", "C"), 
              levels=c("lev1", "lev2"), #levels=c(2,2,2), 
              values=c(0, 0, 2, 3, 1, 2, 1, 4))
class(dat)
```

Network from data: 

```{r}
gr.dG <- compile(grain( dG, data=dat ))
gr.uG <- compile(grain( uG, data=dat ))
```

Extracting information 

```{r}
extractCPT(dat, dG) |> c()
```


```{r}
extractPOT(dat, uG) |> c()
```

